---
title: "Giovanni Scacco - An Italian journey in The United Sates"
---

# Welcome to My Professional Space

## About Me

Hello! I'm **Giovanni Scacco**, an International student from **Italy**, currently pursuing a Master program in Business Analytics at **Baruch College**, specializing in **Accounting**. With a background in **accounting, finance, data analysis, and software asset management**, I thrive at the intersection of technology and business efficiency.

My professional journey has taken me through roles in **accounting, risk advisory, and integration solutions**, including an internship at **Deloitte Risk Advisory**. I am passionate about leveraging data-driven insights to optimize IT governance and financial strategies.

## Let's Connect

📧 [Email Me](mailto:Gio.scacco03@gmail.com)  
🔗 [LinkedIn](linkedin.com/in/giovanni-scacco-2411a1212)  

Stay tuned as I build out this site to share my experiences, projects, and insights!

```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```
