---
title: "Mini-Project 03: Creating the Ultimate Playlist"
author: "Giovanni Scacco"
format: html
editor: visual
---



```{r setup, include=FALSE}
library(tidyverse)
library(jsonlite)
library(stringr)

strip_spotify_prefix <- function(x) {
  str_replace(x, "spotify:(artist|track|album):", "")
}

load_playlists <- function() {
  json_files <- list.files("data/mp03/playlists", pattern = "*.json", full.names = TRUE)

  map_dfr(json_files, function(file) {
    pl_data <- fromJSON(file)
    playlists <- pl_data$playlists

    map_dfr(1:nrow(playlists), function(i) {
      playlist <- playlists[i, ]
      tracks <- playlist$tracks[[1]]

      if (is.null(tracks) || nrow(tracks) == 0) return(NULL)

      tibble(
        playlist_name = rep(playlist$name, nrow(tracks)),
        playlist_id = rep(playlist$pid, nrow(tracks)),
        playlist_followers = rep(playlist$num_followers, nrow(tracks)),
        playlist_position = tracks$pos,
        track_name = tracks$track_name,
        track_uri = tracks$track_uri,
        artist_name = tracks$artist_name,
        artist_uri = tracks$artist_uri,
        album_name = tracks$album_name,
        album_uri = tracks$album_uri,
        duration = tracks$duration_ms
      )
    })
  })
}

# Load and create PLAYLIST_DF
PLAYLIST_DF <- load_playlists() %>%
  mutate(
    track_id = strip_spotify_prefix(track_uri),
    artist_id = strip_spotify_prefix(artist_uri),
    album_id = strip_spotify_prefix(album_uri)
  )

```

---
```{r}
# Load SONGS data
load_songs <- function() {
  url <- "https://raw.githubusercontent.com/gabminamedez/spotify-data/refs/heads/master/data.csv"
  destfile <- "data/mp03/songs/data.csv"

  if (!file.exists(destfile)) {
    dir.create("data/mp03/songs", recursive = TRUE, showWarnings = FALSE)
    download.file(url, destfile, mode = "wb")
  }

  read_csv(destfile)
}

SONGS <- load_songs()

# Load the SONGS dataset
load_songs <- function() {
  url <- "https://raw.githubusercontent.com/gabminamedez/spotify-data/refs/heads/master/data.csv"
  destfile <- "data/mp03/songs/data.csv"

  if (!file.exists(destfile)) {
    dir.create("data/mp03/songs", recursive = TRUE, showWarnings = FALSE)
    download.file(url, destfile, mode = "wb")
  }

  read_csv(destfile)
}

SONGS <- load_songs()


# Exploratory Data Analysis

## Distinct Tracks and Artists

```{r}
PLAYLIST_DF %>% distinct(track_id) %>% nrow()
PLAYLIST_DF %>% distinct(artist_id) %>% nrow()
```

## Most Popular Tracks

```{r}
PLAYLIST_DF %>% 
  group_by(track_name) %>% 
  summarise(appearances = n()) %>% 
  arrange(desc(appearances)) %>% 
  slice_head(n = 5)
```

## Most Danceable Track
SONGS %>% arrange(desc(danceability)) %>% slice(1)


---

# Playlist Curation

Based on anchor songs "One Dance" and "Closer", I explored related songs using playlist co-occurrence, musical features, artist connection, year similarity, and valence matching.

```{r}
ultimate_playlist <- tibble(
  track_order = 1:12,
  track_name = c(
    "Don't Let Me Down",
    "Cold Water",
    "Too Good",
    "Broccoli (feat. Lil Yachty)",
    "Work from Home",
    "Forever",
    "Roses",
    "Party In The U.S.A.",
    "See You Again",
    "Replay",
    "Paper Planes",
    "Soak Up The Sun"
  ),
  artist = c(
    "The Chainsmokers",
    "Major Lazer feat. Justin Bieber & MØ",
    "Drake feat. Rihanna",
    "D.R.A.M. feat. Lil Yachty",
    "Fifth Harmony",
    "Chris Brown",
    "The Chainsmokers",
    "Miley Cyrus",
    "Wiz Khalifa",
    "Iyaz",
    "M.I.A.",
    "Sheryl Crow"
  )
)
```

---

# Playlist Feature Analysis

```{r}
# Enrich Playlist
final_playlist_features <- ultimate_playlist %>%
  left_join(SONGS %>% rename(track_name = name), by = "track_name") %>%
  distinct(track_name, .keep_all = TRUE)

# View playlist with features
final_playlist_features %>%
  select(track_order, track_name, artist, danceability, energy, valence)

```

