---
title: "Mini-Project #04: Exploring Recent US Political Shifts"
author: "Giovanni Scacco"
format: html
editor: visual
---

## Introduction

In this project, I investigated the narrative of a nationwide political shift following the 2024 U.S. Presidential Election. The *New York Times* published a widely-circulated map illustrating a country turning redder, raising questions about whether this shift was overstated.

Acting as a partisan Republican commentator, I aimed to analyze county-level election results to validate a Republican narrative: that 2024 marked a profound political realignment. My work included scraping, cleaning, analyzing, and visualizing election data, as well as developing compelling partisan messaging.

This report documents the process, analysis, and conclusions from this investigation.

---

## Data Collection and Preparation

### Downloading County Shapefiles

I began by downloading U.S. county shapefiles from the U.S. Census Bureau. I ensured the download was automated and cached to avoid redundant downloads.

```{r}
library(sf)
library(dplyr)
library(readr)


data_dir <- "data/mp04"
if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

shapefile_url <- "https://www2.census.gov/geo/tiger/GENZ2022/shp/cb_2022_us_county_5m.zip"
shapefile_zip <- file.path(data_dir, "county_shapefile.zip")

if (!file.exists(shapefile_zip)) {
  download.file(shapefile_url, shapefile_zip)
  unzip(shapefile_zip, exdir = data_dir)
}

shapefile_path <- list.files(data_dir, pattern = "\\.shp$", full.names = TRUE, recursive = TRUE)
counties_sf <- st_read(shapefile_path)
```


I confirmed that r nrow(counties_sf) counties were successfully loaded.

Scraping 2024 and 2020 Election Results
To collect election results, I built a scraping function using httr2 and rvest, iterating over all 50 states. Each state’s results were cached locally to ensure reproducibility and efficiency.

I handled inconsistent column names (e.g., county_61, county_214) and cleaned numeric fields to standardize vote counts across states and years.

```{r}
# --- Load cleaned 2024 election results FIRST ---
results_2024_all_clean <- read_csv("data/mp04/2024_election_results_all_states.csv")

# Define valid county columns
county_cols_2024 <- c("county", "county_214", "county_116", "county_22", "county_23", 
                      "county_156", "county_61", "county_267", "county_31", "county_24", 
                      "county_city", "parish")

# Intersect available columns
valid_county_cols_2024 <- intersect(county_cols_2024, colnames(results_2024_all_clean))

# Create standardized county_name and uppercase state
results_2024_all_clean <- results_2024_all_clean |>
  mutate(
    county_name = coalesce(!!!syms(valid_county_cols_2024)),
    state = toupper(state),
    county_name = toupper(county_name)
  )

# --- Load cleaned 2020 election results ---
results_2020_all_clean <- read_csv("data/mp04/2020_election_results_all_states.csv")

# Define valid county columns for 2020 (you may need to adjust these if different)
county_cols_2020 <- c("county", "county_169", "county_126", "county_31", "county_46", 
                      "county_201", "county_49", "county_33", "county_city", "parish")

valid_county_cols_2020 <- intersect(county_cols_2020, colnames(results_2020_all_clean))

results_2020_all_clean <- results_2020_all_clean |>
  mutate(
    county_name = coalesce(!!!syms(valid_county_cols_2020)),
    state = toupper(state),
    county_name = toupper(county_name)
  )

# --- Clean shapefile ---
counties_sf_clean <- counties_sf |>
  mutate(
    state = toupper(STATE_NAME),
    county_name = toupper(NAME)
  )

# --- Combine election results ---
election_combined <- left_join(
  results_2024_all_clean |> select(state, county_name, trump_2024 = donald_trump_republican, harris_2024 = kamala_harris_democratic, total_2024 = total),
  results_2020_all_clean |> select(state, county_name, trump_2020 = donald_trump_republican, biden_2020 = joe_biden_democratic, total_2020 = total),
  by = c("state", "county_name")
)

# --- Join with shapefile ---
geo_combined <- left_join(counties_sf_clean, election_combined, by = c("state", "county_name"))

# --- Identify county with most Trump votes in 2024 ---
most_trump_votes_2024 <- geo_combined |>
  filter(trump_2024 == max(trump_2024, na.rm = TRUE)) |>
  select(state, county_name, trump_2024)

most_trump_votes_2024
```

➡️ Los Angeles County, California recorded 1,189,862 votes for Trump.

🟢 Counties with the highest Biden vote share in 2020
```qmd
```{r}
```{r}
geo_combined <- geo_combined |>
  mutate(biden_fraction_2020 = biden_2020 / total_2020)

highest_biden_fraction_2020 <- geo_combined |>
  filter(biden_fraction_2020 == max(biden_fraction_2020, na.rm = TRUE)) |>
  select(state, county_name, biden_fraction_2020)

highest_biden_fraction_2020
```

➡️ Oglala Lakota County, South Dakota delivered 88.4% of its votes to Biden.

🟢 Counties with the largest shift toward Trump (absolute)
```{r}
```{r}
geo_combined <- geo_combined |>
  mutate(shift_trump_abs = trump_2024 - trump_2020)

largest_shift_trump <- geo_combined |>
  filter(shift_trump_abs == max(shift_trump_abs, na.rm = TRUE)) |>
  select(state, county_name, shift_trump_abs)

largest_shift_trump
```

➡️ Miami-Dade County, Florida showed an increase of +72,757 votes for Trump between 2020 and 2024.

🟢 State with the smallest shift toward Trump
```{r}
```{r}
state_shift <- geo_combined |>
  group_by(state) |>
  summarise(
    total_trump_2020 = sum(trump_2020, na.rm = TRUE),
    total_trump_2024 = sum(trump_2024, na.rm = TRUE),
    total_harris_2024 = sum(harris_2024, na.rm = TRUE),
    shift_trump = total_trump_2024 - total_trump_2020
  ) |>
  arrange(shift_trump)

state_shift |> slice(1)
```

➡️ Louisiana was the state least shifting toward Trump, with a net change of -47,518 votes.

🟢 Largest county by area
```{r}
```{r}
geo_combined <- geo_combined |>
  mutate(area_sq_meters = as.numeric(st_area(geometry)))

largest_county_area <- geo_combined |>
  filter(area_sq_meters == max(area_sq_meters, na.rm = TRUE)) |>
  select(state, county_name, area_sq_meters)

largest_county_area
```

➡️ Yukon-Koyukuk, Alaska covers 380,497 km², the largest by area.

🟢 County with highest voter density (2020)
```{r}
```{r}
geo_combined <- geo_combined |>
  mutate(area_sq_meters = as.numeric(st_area(geometry)),
         area_sq_km = area_sq_meters / 1e6,
         voter_density_2020 = total_2020 / area_sq_km)

highest_voter_density <- geo_combined |>
  filter(voter_density_2020 == max(voter_density_2020, na.rm = TRUE)) |>
  select(state, county_name, voter_density_2020)

highest_voter_density
```

➡️ Fairfax County, Virginia had the highest density at 37,360 voters per square kilometer.

🟢 County with largest turnout increase in 2024
```{r}
```{r}
geo_combined <- geo_combined |>
  mutate(turnout_change = total_2024 - total_2020)

largest_turnout_increase <- geo_combined |>
  filter(turnout_change == max(turnout_change, na.rm = TRUE)) |>
  select(state, county_name, turnout_change)

largest_turnout_increase
```

➡️ Montgomery County, Texas added +35,715 voters compared to 2020.

Reproducing the NYT Map
I computed the percentage shift in Republican support per county and visualized it using arrows indicating direction and magnitude.

Counties shifting right were colored red; counties shifting left, blue.

```{r}
library(ggplot2)

geo_combined <- geo_combined |>
  mutate(
    pct_trump_2020 = trump_2020 / total_2020,
    pct_trump_2024 = trump_2024 / total_2024,
    shift_pct = pct_trump_2024 - pct_trump_2020
  )

geo_combined_proj <- st_transform(geo_combined, crs = 5070)
centroids_proj <- st_centroid(geo_combined_proj)

arrow_scale <- 50000  # adjust scale for projection

centroids_coords <- centroids_proj |>
  mutate(
    x_start = st_coordinates(geometry)[,1],
    y_start = st_coordinates(geometry)[,2],
    x_end = x_start + arrow_scale * shift_pct,
    y_end = y_start
  )

base_map <- ggplot(geo_combined_proj) +
  geom_sf(aes(fill = shift_pct), color = NA) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Shift Right (%)") +
  theme_void()

centroids_coords_clean <- centroids_coords |> 
  filter(!is.na(x_start), !is.na(x_end), !is.na(shift_pct), abs(shift_pct) >= 0.02)

nyt_map <- base_map +
  geom_segment(
    data = centroids_coords_clean |> st_drop_geometry(),
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end),
    arrow = arrow(length = unit(0.1, "inches")),
    linewidth = 0.2,
    color = "black",
    alpha = 0.4
  )
nyt_map
```

This visualization clearly shows a majority of counties shifting toward the Republican candidate.

Additional Analysis
Statistical Testing: Is the median shift > 5%?
To formally test whether the median county shift exceeded 5%, I used bootstrap resampling via the infer package.

```{r}
set.seed(123)

# manually bootstrap
shift_values <- geo_combined$shift_pct[!is.na(geo_combined$shift_pct)]

bootstrap_medians <- tibble(
  stat = replicate(1000, median(sample(shift_values, replace = TRUE)))
)


# observed median
observed_median <- median(geo_combined$shift_pct, na.rm = TRUE)

# confidence interval
ci <- quantile(bootstrap_medians$stat, probs = c(0.025, 0.975), na.rm = TRUE)

# p-value
p_val <- mean(bootstrap_medians$stat >= observed_median)

ci
p_val

```

✅ 95% Confidence Interval: [1.7%, 1.85%]
✅ p-value: 0.513

Since p > 0.05, I failed to reject the null hypothesis: the median shift was not greater than 5%.

This statistical evidence suggests that, while many counties shifted right, the median shift was modest.

Proportion of Counties Shifting Right
I calculated the proportion of counties experiencing a rightward shift:
```{r}
```{r}
# Calculate proportion of counties shifting right
percent_right_shift <- geo_combined |>
  filter(!is.na(shift_pct)) |>   
  # remove NAs
  summarize(percent = mean(shift_pct > 0) * 100) |>  
  # percent of counties with positive shift
  pull(percent)

percent_right_shift

percent_right_shift
```

➡️ 91.2% of counties shifted right compared to 2020.

Shift Direction Distribution
```{r}
geo_combined |> 
  mutate(shift_dir = ifelse(shift_pct > 0, "Right", "Left")) |>
  count(shift_dir) |>
  ggplot(aes(x = shift_dir, y = n, fill = shift_dir)) +
  geom_col() +
  labs(title = "Direction of Shift by County", y = "Number of Counties") +
  theme_minimal()
```

This figure visually reinforces that most counties shifted right.

Mean Shift in Large Counties
For counties with over 100,000 votes:
```{r}
# Calculate mean shift for counties with over 100,000 votes in 2024
mean_shift_large <- geo_combined |>
  filter(!is.na(shift_pct), total_2024 >= 100000) |>  
  # ensure no NA and population threshold
  summarize(mean_shift = mean(shift_pct)) |> 
  pull(mean_shift)

mean_shift_large

```

➡️ Mean shift among large counties: +1.87%

Partisan Narrative
From a Republican standpoint, the 2024 election signified a broad ideological realignment. Framing the narrative:

**“The red wave is undeniable.
🔴 Over 91% of counties moved toward Trump.
🔴 Even historically blue strongholds like Miami-Dade flipped by +72,000 votes.
🔴 49 states showed increased Republican support; Louisiana stood alone resisting.

Democrats argue ‘land doesn’t vote’—but it’s not about the map. It’s about momentum.
And from every corner of America, voters sent a clear message: America’s moving right.”**

Discussion
Although a partisan narrative emphasizes widespread Republican gains, my statistical findings highlight nuance:

✅ The median county shift was small (1.8%)
✅ Large urban counties barely shifted (mean +1.9%)
✅ The visual impact of red-dominant maps belies population-weighted trends

The Republican narrative is visually compelling but statistically restrained.

This underscores the power of geographic visualization in shaping perceptions—maps with more “red” don’t always reflect voter counts, but they do influence narratives.

Conclusion
In this project, I scraped, cleaned, and analyzed 2024 and 2020 election data at the county level. I reproduced the NYT map and conducted inferential tests to assess the narrative of a national rightward shift.

While the majority of counties shifted toward Trump, the magnitude of those shifts was modest in population centers. The narrative of a sweeping ideological transformation holds in a geographic sense, but less so in vote-weighted terms.

This analysis demonstrates how data can be wielded to craft persuasive partisan messages—even when the underlying statistics are more nuanced.

References
Wikipedia: “2024 United States Presidential Election in [State]”

U.S. Census Bureau: TIGER/Line Shapefiles


